Diagnostic Potential & the Interactive Dynamics of the Colorectal Cancer Virome

### 2018-07-08 doc/manuscript.Rmd
Pushing a couple of the edits.

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the bacteriophage signal may be a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal. If this hypothesis was true, we would expect a correlation between the relative abundances of influential bacteria and phages. Instead, we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**. Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages were not primarily reflections of influential bacteria, and suggest a more complex relationship between the bacterial and phage communities.

@@

Using our findings here, we have begun a path toward the delineation of the role the colonic virome plays in colorectal cancer in the form of a working hypothesis that will guide our future studies **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phages indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

@@

At this time, our working hypothesis represents a conceptualization of areas for future work, which will be required to characterize the colorectal cancer microbiome at the functional, mechanistic level. There are many different ways in which this system can be operating, and our working hypothesis is one. For example, it is possible that the bacterial communities cause a change in the virome instead of the virome altering the bacterial communities. Reviewing every possible mechanism and characteristic of the system is beyond the scope of this manuscript, but the results we present provide a step toward the answer. Important future directions will include larger cohort human studies, further *in vitro* and *in vivo* mechanistic experimentation, and even attempts at community studies using absolute abundance values instead of relative abundance, which would allow for more accurate community dynamic modeling.

As with most microbiome studies it is important to note that, despite including data for both bacteria and viruses, we are still only studying a subset of the microbiome. Here we are building on the bacterial-centric literature around colorectal cancer by beginning to characterize the virome. The caveat here is that we are technically limited to studying only the DNA virome, and therefore are not measuring RNA viruses. This will be an important direction for future studies, as well as beginning to characterize other additional microbial communities such as fungi, archaea, etc.

@@

![*Basic Quality Control Metrics. A) VLP genomic DNA yield from all sequenced samples. Each bar represents a sample which is grouped and colored by its associated disease group. B) Sequence yield following quality control including quality score filtering and human decontamination. Dashed line shows rarefaction line in which all samples with lower sequence yield were excluded from downstream analysis.*\label{qualcontrol}]()

@@

![*Beta-diversity comparing disease states and the study negative controls which remained following rarefaction. Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test. Comparisons in which the intervals cross the zero mean difference line (dashed line) were not significantly different.*\label{betaogunegative}]()

>>>>

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we hypothesized that the bacteriophage signal may be a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal. If this hypothesis was true, we would expect a correlation between the relative abundances of influential bacteria and phages. Instead, we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**. Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages were not primarily reflections of influential bacteria, and suggests a more complex relationship between the bacterial and phage communities.

@@

Based on our findings, we have developed a conceptual model to be tested in future studies aimed at elucidating the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a machine learning approach (random forest classification) to leverage inherent, complex patterns not detected by diversity measures, we were able to detect strong associations between the colon virus community composition and colorectal cancer. The dsDNA virome of colorectal cancer was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phages indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

@@

Our working hypothesis represents a conceptualization of areas for future work, which will be required to characterize the colorectal cancer microbiome at the functional, mechanistic level. There are many different ways in which this system can be operating, and our working hypothesis is one. For example, it is possible that the bacterial communities cause a change in the virome instead of the virome altering the bacterial communities. Reviewing every possible mechanism and characteristic of the system is beyond the scope of this manuscript, but the results we present provide a step toward the answer. Important future directions will include larger cohort human studies, further *in vitro* and *in vivo* mechanistic experimentation, and even attempts at community studies using absolute abundance values instead of relative abundance, which would allow for more accurate community dynamic modeling.

@@

![*Basic Quality Control Metrics. A) VLP genomic DNA yield from all sequenced samples. Each bar represents a sample which is grouped and colored by its associated disease group or no-DNA negative control. B) Sequence yield following quality control including quality score filtering and human decontamination. Dashed line indicates rarefaction level ($10^6$ reads) in which all samples with lower sequence yields less than this level were excluded from downstream analysis.*\label{qualcontrol}]()

@@

![*Beta-diversity analysis comparing Bray-Curtis dissimilarity between disease state and negative control community structures that were captured following sequence rarefaction. Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test. Comparisons in which the intervals cross the zero mean difference line (dashed line) were not significantly different.*\label{betaogunegative}]()
